Estimating the Initial Susceptible Fraction in Dengue Dynamics

### 2015-06-15 draft_lm1.tex
defined S_0 in equation \label{eq:effbeta}; closes #33
resolving confusing notation; fixes #31

We explore the approach of Ederer and Mantel~\citep{mantel}, whose objective is 
to obtain 
confidence intervals for the ratio of two Poisson counts. 
Let $Y_{t} \sim Poisson(\lambda_t)$ and $Y_{t+1} \sim Poisson(\lambda_{t+1})$ 
and define $S = Y_{t} + Y_{t+1}$.
The authors note that by conditioning on the sum $S$
\begin{align}
\label{eq:binlike}
Y_{t+1} | S &\sim Binomial(S, \theta_t) \\
\theta_t &= \frac{\lambda_{t+1}}{\lambda_{t} + \lambda_{t+1}}
\end{align}
Let $c_{\alpha}(\theta_t) = \{\theta_t^{(L)} , \theta_t^{(U)} \}$ be such that 
$Pr(\theta_t^{(L)}<\theta_t <\theta_t^{(U)}) = \alpha$.
Analogously, define $c_{\alpha}(R_t) = \{R_t^{(L)} , R_t^{(U)} \}$ such that 
$Pr(R_t^{(L)}<R_t<R_t^{(U)}) = \alpha$.
Ederer and Mantel (1974)~\citep{mantel} show that one can construct a 
$100\alpha 
\%$ confidence interval for \rr by noting that
\begin{equation}
\label{eq:confRt}
 R_t^{(L)} = \frac{\theta_t^{(L)}}{(1-\theta_t^{(L)})} \quad \text{and} \quad 
R_t^{(U)} = \frac{\theta_t^{(U)}}{(1-\theta_t^{(U)})}\\
\end{equation}
Because the transform from $\theta$ to \rr is monotonically 
increasing, the result holds for confidence and credibility intervals alike.

Many authors have chosen to quantify the uncertainty about $\theta$ 
following orthodox approaches  (see for example~\citep{wilson} 
and~\citep{clopper}) mainly for simplicity.
We choose instead to take a Bayesian approach and use the  $100\alpha \%$ 
posterior credibility interval for $\theta_t$ as $c_{\alpha}(\theta_t)$.
If we choose the conjugate beta prior with parameters $a_0$ and $b_0$ for the 
binomial likelihood in (\ref{eq:binlike}), the posterior distribution for 
$\theta_t$ is
\begin{equation}
\label{eq:thetapost}
p(\theta_t| Y_{t+1}, S) \sim Beta(Y_{t+1} + a_0, Y_t + b_0)
\end{equation}
Combining equations~(\ref{eq:confRt}) and~(\ref{eq:thetapost}) 
tells us that the induced posterior distribution of $R_t$ is 
a beta prime (or inverted beta) with parameters $ a_1 = Y_{t+1} + a_0$ and $b_1 
=  Y_t + b_0$~\citep{dubey1970}.
The density of the induced distribution is then 
\begin{equation}
\label{eq:densityMantel}
f_P(R_t| a_1, b_1) = \frac{\Gamma(a_1 + b_1)}{\Gamma(a_1)\Gamma(b_1)} R_t^{a_1 
- 
1} (1 + R_t)^{-(a_1 + b_1)}
\end{equation}
Thus, the expectation of \rr is $a_1/(b_1 - 1)$ and its variance is 
$a_1(a_1 + b_1 - 1)/\left((b_1 - 2)(b_1 - 1)^2 \right) $.
Note that this result holds only for $n = 1$.
Sampling from the posterior in~(\ref{eq:densityMantel}) can be made 
straightforward by first sampling from~(\ref{eq:thetapost}) and then applying 
the transform in~(\ref{eq:confRt}).
Also, one can choose $a_0$ and $b_0$ so as to elicit meaningful prior 
distributions for $R_t$.
We show how to elicit the prior for $R_t$ from specified prior mean and variance
or coefficient of variation in the \textbf{Appendix}.

@@

For our analysis we chose to let the force of infection vary with time, just 
as it does in the actual epidemics, as seen in the data. So as the epidemic 
progresses, the effective transmission  rate changes and is 
given by 
\begin{equation} 
 \label{eq:effbeta}
 \beta(t) = \frac{R_t\cdot\tau}{S_0}
\end{equation}
where $R_t$ is the effective reproductive number, estimated as 
in~\ref{eq:Rtestimate}.
The complete model with the time-varying force of infection is given by
the system of ordinary differential equations:
\begin{align}
   \label{eq:model}
 \frac{dS}{dt} &= -\beta(t)SI \\     \nonumber
 \frac{dI}{dt} &= \beta(t)SI - \tau I&\\      \nonumber
 \frac{dR}{dt} &= \tau I&
\end{align}  
where $S + I + R = 1 \: \forall\: t$. % \in [T_0, T_1]$. 
Of course, this is a rather simplified model, in which, for instance, the 
vector is omitted.
The rationale for this simplification is based on the ability of the 
empirically derived $R_t$  to incorporate the effects of the fluctuating vector 
populations.
Also, although there are multiple circulating serotypes, our approach
can not discriminate between them due to the lack of serotype-specific data.
Nevertheless, this modelling strategy can still provide some insight into the 
disease dynamics and allows us to estimate the initial fraction of susceptibles 
$S_0$, a key epidemiological parameter.

>>>>

We explore the approach of Ederer and Mantel~\citep{mantel}, whose objective is 
to obtain 
confidence intervals for the ratio of two Poisson counts. 
Let $Y_{t} \sim Poisson(\lambda_t)$ and $Y_{t+1} \sim Poisson(\lambda_{t+1})$ 
and define $W = Y_{t} + Y_{t+1}$.
The authors note that by conditioning on the sum $W$
\begin{align}
\label{eq:binlike}
Y_{t+1} | S &\sim Binomial(W, \theta_t) \\
\theta_t &= \frac{\lambda_{t+1}}{\lambda_{t} + \lambda_{t+1}}
\end{align}
Let $c_{\alpha}(\theta_t) = \{\theta_t^{(L)} , \theta_t^{(U)} \}$ be such that 
$Pr(\theta_t^{(L)}<\theta_t <\theta_t^{(U)}) = \alpha$.
Analogously, define $c_{\alpha}(R_t) = \{R_t^{(L)} , R_t^{(U)} \}$ such that 
$Pr(R_t^{(L)}<R_t<R_t^{(U)}) = \alpha$.
Ederer and Mantel (1974)~\citep{mantel} show that one can construct a 
$100\alpha 
\%$ confidence interval for \rr by noting that
\begin{equation}
\label{eq:confRt}
 R_t^{(L)} = \frac{\theta_t^{(L)}}{(1-\theta_t^{(L)})} \quad \text{and} \quad 
R_t^{(U)} = \frac{\theta_t^{(U)}}{(1-\theta_t^{(U)})}\\
\end{equation}
Because the transform from $\theta$ to \rr is monotonically 
increasing, the result holds for confidence and credibility intervals alike.

Many authors have chosen to quantify the uncertainty about $\theta$ 
following orthodox approaches  (see for example~\citep{wilson} 
and~\citep{clopper}) mainly for simplicity.
We choose instead to take a Bayesian approach and use the  $100\alpha \%$ 
posterior credibility interval for $\theta_t$ as $c_{\alpha}(\theta_t)$.
If we choose the conjugate beta prior with parameters $a_0$ and $b_0$ for the 
binomial likelihood in (\ref{eq:binlike}), the posterior distribution for 
$\theta_t$ is
\begin{equation}
\label{eq:thetapost}
p(\theta_t| Y_{t+1}, W) \sim Beta(Y_{t+1} + a_0, Y_t + b_0)
\end{equation}
Combining equations~(\ref{eq:confRt}) and~(\ref{eq:thetapost}) 
tells us that the induced posterior distribution of $R_t$ is 
a beta prime (or inverted beta) with parameters $a_1 = Y_{t+1} + a_0$ and $b_1 
=  Y_t + b_0$~\citep{dubey1970}.
The density of the induced distribution is then 
\begin{equation}
\label{eq:densityMantel}
f_P(R_t| a_1, b_1) = \frac{\Gamma(a_1 + b_1)}{\Gamma(a_1)\Gamma(b_1)} R_t^{a_1 
- 
1} (1 + R_t)^{-(a_1 + b_1)}
\end{equation}
Thus, the expectation of \rr is $a_1/(b_1 - 1)$ and its variance is 
$a_1(a_1 + b_1 - 1)/\left((b_1 - 2)(b_1 - 1)^2 \right) $.
Note that this result holds only for $n = 1$.
Sampling from the posterior in~(\ref{eq:densityMantel}) can be made 
straightforward by first sampling from~(\ref{eq:thetapost}) and then applying 
the transform in~(\ref{eq:confRt}).
Also, one can choose $a_0$ and $b_0$ so as to elicit meaningful prior 
distributions for $R_t$.
We show how to elicit the prior for $R_t$ from specified prior mean and variance
or coefficient of variation in the \textbf{Appendix}.

@@

For our analysis we chose to let the force of infection vary with time, just 
as it does in the actual epidemics, as seen in the data. So as the epidemic 
progresses, the effective transmission  rate changes and is 
given by 
\begin{equation} 
 \label{eq:effbeta}
 \beta(t) = \frac{R_t\cdot\tau}{S_0}
\end{equation}
where $R_t$ is the effective reproductive number, estimated as 
in~\ref{eq:Rtestimate} and $S_0$ is the initial fraction of susceptible 
individuals.
The complete model with the time-varying force of infection is given by
the system of ordinary differential equations:
\begin{align}
   \label{eq:model}
 \frac{dS}{dt} &= -\beta(t)SI \\     \nonumber
 \frac{dI}{dt} &= \beta(t)SI - \tau I&\\      \nonumber
 \frac{dR}{dt} &= \tau I&
\end{align}  
where $S + I + R = 1 \: \forall\: t$. % \in [T_0, T_1]$. 
Of course, this is a rather simplified model, in which, for instance, the 
vector is omitted.
The rationale for this simplification is based on the ability of the 
empirically derived $R_t$  to incorporate the effects of the fluctuating vector 
populations.
Also, although there are multiple circulating serotypes, our approach
can not discriminate between them due to the lack of serotype-specific data.
Nevertheless, this modelling strategy can still provide some insight into the 
disease dynamics and allows us to estimate the initial fraction of susceptibles 
$S_0$, a key epidemiological parameter.
